Probabilistic Programming

### 2014-04-21 paper/paper.tex
Reduce margins
Add Sample AST
Add Source link
(Mostly) finish paper

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}   
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}

\usepackage{qtree}
\usepackage{tikz}
\usepackage{tikz-qtree}

\usepackage[]{algorithm2e}
\hyphenation{ECMAScript JavaScript escodegen esprima}

@@

An abstract syntax tree (AST) is a tree-based, rather than text-based, representation of a program.
Such a tree is generated using the grammar rules of a programming language and text input. In the tree, \emph{nodes}
are language control structures such as {\tt for} loops, {\tt if} statements, variable and function
declarations, and so on. The \emph{edges} connecting a parent node to its children represent the different \emph{features}
of that node. For instance, a {\tt for} loop would have the features \emph{init} (e.g. {\tt var i = 0;}),
\emph{test} (e.g. {\tt i < array.length;}), \emph{update} (e.g. {\tt i++}), and a \emph{body} (a list
of statements that should be executed per loop iteration). To implement lists, we include the additional
\emph{follow} edge as a feature to all statements, which would connect a statement node to the next statement
in the list. If the statement is last in its container list, we connect its \emph{follow} edge to a special
{\tt {\_end}} node.

\begin{center}

[example of AST here]

@@

\end{center}

Our system deals with programs at the level of AST's as they provide more structure and modularity than text.
We use the syntax tree format of Mozilla's Parser API \cite{parser api}, and \emph{esprima} \cite{esprima}, an ECMAScript
parser, to transform program source code into an AST of this format.

%escodegen \cite{escodegen} (ECMAScript code generator) to transform ASTs into program source code.

@@

To build our Markov chain structure, we parse a corpus of JavaScript files into their
respective AST's, and then collect statistics from these trees to build our Markov model.

The model maps AST paths to the probability distributions of node types or values for that given
path. An AST path is a list of items alternating between a node type and one of its features, thus encoding the unique position of a 
specific node in the AST. The probability distribution for an AST path is the set of possible node types or values that would appear in the
syntax tree following the given path, and the respective probabilities of their
occurance following the given path.

More formally, given a depth $d$ (indicating how many nodes we should backtrace) for a given
${node}_d$ and feature ${edge}_d$, the generation of ${node}_d$'s child node given by the
transition ${edge}_d$, can be expressed by the following function $f$, \begin{align*}
f({node}_1, {edge}_1, ..., {node}_{d}, {edge}_{d}) = \begin{dcases*}
		{newnode_1},	&	$P({newnode_1})$ \\
		{newnode_2},	&	$P({newnode_2})$ \\
		...\\
		{newnode_n},    &   $P({newnode_n})$ \\
	\end{dcases*}
\end{align*}

such that ${Path}({node}_1, {edge}_1) = {node}_2$, ${Path}({node}_2, {edge}_2) = {node}_3$, ..., and ${Path}({node}_d, {edge}_d) = {newnode}_k$, where ${newnode}_k$ is randomly selected from ${newnode}_1, ..., {newnode}_k$ given the probabilities $P({newnode_1}), ..., P({newnode_n})$, and \begin{align*}
\sum\limits_{i = 0}^n P({newnode_i}) = 1.
\end{align*}

From this it can be seen that generating programs is simply a repetitive application of this function, where for the chosen ${newnode}_k$, we can generate each child of ${newnode}_k$ with, $\forall {edge} \in {Features}({newnode}_k)$, \begin{align*}
{Path}({newnode}_k, {edge}) = f({node}_2, {edge}_2, ..., {node}_{d}, {edge}_{d}, {newnode}_k, {edge}).
\end{align*}

This process will eventually terminate as the probability of generating {\tt \_end} nodes for a \emph{follow} edge will always be non-zero -- as the programs we parse into our corpus are surely finite -- and such nodes have no features to be ``expanded.''

@@

Building the Markov chain model means being able to provide the probability distribution of new nodes for arbitrary input,
as described by the function in Section 2.2. To do this, we constructed a data structure using JavaScript Object Notation (JSON),
a structural format useful for ease of specification and integration with JavaScript programs.

The data structure is a hash table (also commonly called a map, dictionary, or associative array), that, on a given input, returns another hash table for paths that begin with that input. This behavior is repeated until the number of inputs reaches the desired depth, at which point a final hash table, mapping the node type to its probability distribution is returned. This, in effect, is modularizing the function $f$ described above, so that we can reach the probability distribution with minimal code and data duplication. This is demonstrated by following example, for  ${depth} = 2$. \begin{align*}
lookup(lookup(lookup(lookup({model}, {node}_1), {edge}_1), {node}_2), {edge}_2) = \textbf{P}({nodes}).
\end{align*}

@@

\begin{verbatim}
procedure addToModel(node, path):
    model = corpusModel
    for each elem in path[path.length-depth*2 .. path.length]:
        model = model[elem]
    model[node.type] += 1
    model[_total] += 1
    for each feature of node:
        temp_path = path + [node.type, feature]
        if node.feature is list of nodes:
            for each item in node.feature:
                addToModel(item, temp_path)
                temp_path += [item.type, "FOLLOW"]
            addToModel(END_NODE, temp_path)
        else:
            addToModel(node.feature, temp_path)

@@

When this process is complete, the JSON object is saved to a file and can be augmented by more code samples thereafter. (An example of such a model is shown in the Appendix.)

See figure~\ref{fig:sample-model} in the appendix for a model for the program
``var x = 34;" In this model we see for example that the path
\texttt{[Program, body, VariableDeclaration, declarations]} leads to
\texttt{VariableDeclarator} with probability 1. This indicates that a Program
node whose body is a VariableDeclaration always has a declaration that is of
type VariableDeclarator.

The model is implemented as a JSON object where the path is indexed into the
object by successively indexing by each string of the path.

@@

\subsubsection{Random programs}

Random programs can be generated rather easily using the approach explained in Section 2.2. The program tree is built up one node at a time, each one generated based on its unique AST path. We always begin a program with a ``Program'' node

\subsubsection{Snippets}

% application of markov chain to abstact syntrax tree

@@

\subsection{vim autocomplete script}

@@

@@

\begin{figure}[h!]
	\caption{model-sample.json}
	\label{fig:sample-model}
	\centering
	A model generated from the program ``var x = 34;"
\end{figure}

\lstinputlisting{model-sample.json}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}   
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage[margin=3cm]{geometry}

%\usepackage{qtree}
%\usepackage{tikz}
%\usepackage{tikz-qtree}

%\usepackage[]{algorithm2e}
\hyphenation{ECMAScript JavaScript escodegen esprima}

@@

An abstract syntax tree (AST) is a tree-based, rather than text-based, representation of a program.
Such a tree is generated using the grammar rules of a programming language and text input. In the tree, \emph{nodes}
are language control structures such as {\tt for} loops, {\tt if} statements, variable and function
declarations, and so on. The \emph{edges} connecting a parent node to its children represent the different \emph{features}
of that node. For instance, a {\tt for} loop would have the features \emph{init} (e.g. {\tt var i = 0;}),
\emph{test} (e.g. {\tt i < array.length;}), \emph{update} (e.g. {\tt i++}), and a \emph{body} (a list
of statements that should be executed per loop iteration). To implement lists, we include the additional
\emph{follow} edge as a feature to all statements, which would connect a statement node to the next statement
in the list. If the statement is last in its container list, we connect its \emph{follow} edge to a special
{\tt {\_end}} node. (See figure~\ref{fig:sample-ast} for an example of a AST.)

% \begin{center}

@@

% \end{center}

Our system deals with programs at the level of AST's as they provide more structure and modularity than text.
We use the syntax tree format of Mozilla's Parser API \cite{parser api}, and \emph{esprima} \cite{esprima}, an ECMAScript
(the language standard of which JavaScript is an implementation)
parser, to transform program source code into an AST of this format.

@@

To build our Markov chain structure, we parse a corpus of JavaScript files into their
respective AST's, and then collect statistics from these trees to establish probability distributions of nodes for a given AST context.

An \emph{AST path} is a list of items, alternating between node types and features, that encodes the unique position of a 
specific node in the AST. The \emph{probability distribution} for an AST path is the set of possible node types or values that would appear in the
syntax tree following the given path, and the respective probabilities of their
occurrence at that point. The model maps AST paths to the probability distributions of node types (or values, as we'll see later) for that given
path.

More formally, given a depth $d$ (indicating how many nodes we should backtrace) for a given
${node}_d$ and feature ${edge}_d$, the generation of ${node}_d$'s child node given by
 ${edge}_d$ can be expressed by the following function $f$: \begin{align*}
f({node}_1, {edge}_1, ..., {node}_{d}, {edge}_{d}) = \begin{dcases*}
		{newnode_1},	&	$P({newnode_1})$ \\
		{newnode_2},	&	$P({newnode_2})$ \\
		...\\
		{newnode_n},    &   $P({newnode_n})$ \\
	\end{dcases*},
\end{align*}

where \begin{align*}
\sum\limits_{i = 0}^n P({newnode_i}) = 1.
\end{align*}

and ${Path}({node}_1, {edge}_1) = {node}_2$, ${Path}({node}_2, {edge}_2) = {node}_3$, ..., ${Path}({node}_d, {edge}_d) = {newnode}_k$, where ${newnode}_k$ is randomly selected from ${newnode}_1, ..., {newnode}_n$ given the probabilities \break $P({newnode_1}), ..., P({newnode_n})$. From this it can be seen that generating programs is simply a repetitive application of this function, where for the chosen ${newnode}_k$, we can generate each child of ${newnode}_k$ with the following formula. \begin{align*} &\forall {edge} \in {Features}({newnode}_k), \\
& \hspace{3em} {Path}({newnode}_k, {edge}) = f({node}_2, {edge}_2, ..., {node}_{d}, {edge}_{d}, {newnode}_k, {edge}).
\end{align*}

This process will eventually terminate as the probability of generating {\tt \_end} nodes for a \emph{follow} edge will always be non-zero -- the programs we parse into our corpus are surely finite -- and such nodes have no features to be ``expanded.''

@@

Building the Markov chain model means being able to provide the probability distribution of new nodes for arbitrary input,
as described by the function in Section 2.2. To do this, we construct a data structure using JavaScript Object Notation (JSON),
a structural format useful for ease of specification and integration with JavaScript programs.

The data structure is a hash table (also commonly called a map, dictionary, or associative array), that, on a given input, returns another hash table for paths that begin with that input. This behavior is repeated until the number of inputs reaches the desired depth, at which point a final hash table, mapping the node type to its probability distribution, is returned. This, in effect, is modularizing the function $f$ described above, so that we can reach the probability distribution with minimal code and data duplication. This is demonstrated by following example, given  ${depth} = 2$. \begin{align*}
lookup(lookup(lookup(lookup({Model}, {node}_1), {edge}_1), {node}_2), {edge}_2) = \textbf{P}({newnodes}).
\end{align*}

@@

\begin{verbatim}
procedure addToModel(node, path):
    model = CorpusModel
    for each elem in path[path.length-depth*2 .. path.length]:
        model = model[elem]
    model[node.type] += 1
    model[_total] += 1
    for each feature of node:
        temp_path = path + [node.type, feature]
        if node.feature is list of nodes:
            for each item in node.feature:
                addToModel(item, temp_path)
                temp_path += [item.type, "FOLLOW"]
            addToModel(END_NODE, temp_path)
        else:
            addToModel(node.feature, temp_path)

@@

When this process is complete, the JSON object is saved to a file and can be augmented by more code samples thereafter. (An example of such a model is shown in figure~\ref{fig:sample-model} in the Appendix. In this model we see, for instance, that the path
[\emph{Program, body, VariableDeclaration, declarations}] leads to
\emph{VariableDeclarator} with probability 1. This indicates that a \emph{Program}
node whose \emph{body} is a \emph{VariableDeclaration} always has a declaration that is of
type \emph{VariableDeclarator}.)

@@

Code can be generated rather easily using the approach explained in Section 2.2. The program tree is built up one node at a time, each one generated based on its unique AST path. The following algorithm demonstrates this process.

\begin{verbatim}
procedure generateNode(model, path):
    for each elem in path[path.length-depth*2 .. path.length]:
        model = model[elem]
    nodeType = pickRandomNodeType(model)
    node = {`type':nodeType}
    for each feature of nodeType:
        new_path = path + [node.type, feature]
        if feature is a list feature:
            nodes = []
            while (newNode = generateNode(model, new_path) != END_NODE:
                nodes += [nodeNode]
                new_path += [newNode.type, "FOLLOW"]
            node.feature = nodes
        else:
            node.feature = generateNode(model, new_path)
    return node

generateNode(CorpusModel, [])
\end{verbatim}

The algorithm looks very similar to the parse algorithm, as it is almost the same process, simply inverted. Rather than examine nodes, determine a path by building it up, and then increment a count in the model, this algorithm first finds the counts in the model for a \emph{given} path, generates a node, and generates nodes for all of that node's features recursively. The result is a generated AST for our program, which is then fed into \emph{escodegen} \cite{escodegen}, an ECMAScript code generator to output program source code that we can display and run.

To generate random programs, we always begin with a ``Program'' node, and then add nodes to its \emph{body} feature (an array.) However, we quickly found that while entertaining, generating random programs had little potential for real application. Instead, we turned to generating code snippets (to be placed in specific contexts of the syntax tree.) Since our algorithm was general enough for this purpose, no more had to be done besides knowing the path for the node we wanted to generate.

\subsubsection{Identifiers and literals}

Naturally, identifiers and literals are a very large part of programming. Without them, code would be entirely unspecific, and would only consist of syntactic symbols and keywords. Yet they pose an issue with code generation, as we found generating identifiers or literals that are likely to be what the user is expecting an especially difficult problem to solve, let alone model. Struggling to come up with a simple, intelligent, and robust solution, and short on time, we fell back to using our existing Markov approach.

In the current implementation of this system, following the paths [\emph{Identifier}, \emph{Name}], or [\emph{Literal}, \emph{Raw}] result not in node types, but rather in the text of the identifier names or literal values. This way, generating identifiers and literals is simply a probabilistic process like the rest of the code generation. Of course, this means that sometimes identifiers are added that are not in scope, that have mismatched types, or that simply make no semantic sense. (But adjusting $depth$ did seem to make programs more meaningful, expectedly. This can be seen experimentally on our web IDE, explained in Section 4.1.)

@@

\subsection{\emph{vim} autocomplete script}

@@

\subsection{Source}

The source code for Probabilistic Programming is available online under an open
source license:\\
\url{https://github.com/RocHack/jschain}.\\
The web IDE can be found here:
\url{https://csug.rochester.edu/u/clehner/jschain/web/}.

@@

\begin{figure}[h!]
	\caption{ast-sample.json}
	\label{fig:sample-ast}
	\centering
	An AST generated from the program ``var x = 34;"

	\lstinputlisting{ast-sample.json}
\end{figure}

\begin{figure}[h!]
	\caption{model-sample.json}
	\label{fig:sample-model}
	\centering
	A model generated from the program ``var x = 34;"

	\lstinputlisting{model-sample.json}
\end{figure}
